Scikit-learn: API design for machine learning

### 2013-06-19 paper.tex
copyedit, trim
Three substantive changes:
* the implementation section no longer discusses Python limitations;
* transformers have fit_transform, not all estimators;
* vectorizers don't only handle text documents.

Started in 2007, \sklearn is developed by an international team of a dozen of
core developers, mostly researchers from various fields (e.g., machine learning,
natural language processing, or neuroscience). The project also greatly benefits
from many occasional contributors proposing small code or documentation fixes or
improvements. Development proceeds on GitHub\footnote{\url{https://github.com/scikit-learn}},
a platform which greatly facilitates this kind of
collaboration. Because of the large number of developers, emphasis is
put on keeping the project maintainable. In particular, code must follow
specific quality guidelines, such as style consistency and unit-test coverage.
In addition, complete documentation and examples are required for all features.
Finally, additions to the code base must pass code review by at least two
developers not involved in the implementation of the proposed addition.

Within the past few years, the project has proved valuable to an ever increasing
user base. Its popularity can be gaged from various indicators such as hundreds
of citations in scientific publications, successes in various machine learning
challenges (e.g., Kaggle), or significant statistics derived from our
repositories and mailing lists.  At the time of writing, the project is watched
by 1309 people and forked 667 times on GitHub, and the mailing-list receives
more than 300 mails per month. The version control logs for the development
version
% ddaa494c116e3c16bf032003c5cccbed851733d2
also count 183 unique contributors to the project.

In a previous paper \citep{pedregosa2011}, \sklearn was briefly presented and
benchmarked against several competitors. In this paper, we instead focus on an
in-depth discussion and analysis of the design choices used in building the
library. We first present in section \ref{sec:core-api} the central application
programming interface (API) and then describe, in section \ref{sec:advanced-api},
advanced API mechanisms built from the core interface. Section
\ref{sec:implementation} describes implementation principles and discusses
limitations of the technologies we rely on. Section \ref{sec:comparison} then
compares \emph{scikit-learn} to other major projects in terms of API. Finally,
conclusions and directions of future works are included in section
\ref{sec:conclusions}.

@@

As an example to illustrate the initialize-\texttt{fit} sequence,
let us consider a supervised learning task using logistic regression.
Given the API defined above, solving this problem is as simple as:\footnote{
  \texttt{import} statement are omitted for brevity
  throughout this paper.}
\begin{pythoncode}
clf = LogisticRegression(penalty="l1")
clf.fit(X_train, y_train)
\end{pythoncode}
In this snippet, a \texttt{LogisticRegression} estimator is first initialized by
setting the \texttt{penalty} hyper-parameter to \texttt{"l1"}, hence enabling
$\ell_1$ regularization. Other hyper-parameters (such as \texttt{C}, which
defines the strength of the regularization) are not explicitly defined and are
thus set to the default values. Upon calling \texttt{fit}, the model is then
learned from the training arrays \texttt{X\_train} and \texttt{y\_train}.
(In this example, \texttt{fit} is called solely for its side effects.)
Since all estimators share the same interface, using a different learning algorithm is
as simple as replacing the constructor (the class name).
For example, to build a random forest on
the same training data, one would simply have to replace
\texttt{LogisticRegression(penalty="l1")} in the snippet above by \\
\texttt{RandomForestClassifier()}.

@@

Since it is common to modify or filter data before feeding it to a learning
algorithm, some estimators in the library implement a \textit{transformer}
interface which defines a \texttt{transform} method. It takes as input some new
data \texttt{X\_test} and yields as output a transformed version of
\texttt{X\_test}. Preprocessing, feature selection and dimensionality reduction
algorithms are all provided as transformers within the library.  In our example,
if one wanted to preprocess the inputs \texttt{X\_train} before fitting the
logistic regression estimator, e.g., by standardizing the feature values, one
could thus simply write:
\begin{pythoncode}
scaler = StandardScaler()
scaler.fit(X_train)
X_train = scaler.transform(X_train)
\end{pythoncode}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}.
Since a \texttt{StandardScaler} stores the mean and standard deviation
that it computed for the training set,
applying it to an unseen test set \texttt{X\_test}
maps that into the appropriate region of feature space:
\begin{pythoncode}
X_test = scaler.transform(X_test)
\end{pythoncode}
Transformers also include a variety of learning algorithms, such as
dimension reduction (PCA, manifold learning techniques), kernel
approximation techniques, and any other approach that can be used to
derive a modified representation of the data.

@@

Furthermore, every estimator allows \texttt{fit(X\_train).transform(X\_train)}
to be written as \texttt{fit\_transform(X\_train)}.
The combined \texttt{fit\_transform} method prevents repeated computations.
Depending on the transformer,
this may either skip only an input validation step,
or in fact use a more efficient algorithm for the transformation.
In the same spirit, clustering algorithms provide a
\texttt{fit\_predict} method
that is equivalent to \texttt{fit} followed by \texttt{predict},
and thus returns the cluster labels assigned to the training data samples.

@@

As an example, a logistic regression classifier
uses by default a one-vs.-rest scheme
for performing multiclass classification.
If a one-vs.-one scheme is desired instead,
this can achieved using a \texttt{OneVsOneClassifier} meta-estimator
wrapping a \texttt{LogisticRegression} estimator:
\begin{pythoncode}
ovo_lr = OneVsOneClassifier(LogisticRegression(penalty="l1"))
\end{pythoncode}
The \texttt{OneVsOneClassifier} object
\textit{clones} the LR estimator multiple times for learning,
resulting in a set of $\frac{K(K-1)}{2}$ estimator objects
for $K$-way classification,
all with the same settings,
that each perform a binary classification and vote to make the final decision.
The snippet exemplifies the importance
of separating object instantiation and actual learning.

@@

\texttt{Pipeline} objects are used for chaining multiple estimators into a
single one. This is useful since a machine learning workflow typically involves
a fixed sequence of steps in processing data (e.g., feature extraction,
normalization, feature selection, learning and making predictions). A sequence
of $N$ such steps can be combined into a composite object called a
\texttt{Pipeline}. The first $N-1$ steps must be transformers and the last can
be either a predictor or a transformer. Conceptually, fitting a pipeline to a
training set amounts to the following recursive procedure: i) when only one step
remains, call its \texttt{fit} method; ii) otherwise, \texttt{fit} the first
step, use it to \texttt{transform} the training set and \texttt{fit} the rest of
the pipeline with the transformed data. The pipeline exposes all the methods the
last estimator in the pipe exposes. That is, if the last estimator is a
predictor, the pipeline can itself be used as a predictor. If the last estimator
is a transformer, then the pipeline is itself a transformer.

A \texttt{FeatureUnion} object is a transformer combining multiple transformers
into a single one that concatenates their outputs.
I.e., a union of transformers that map inputs having $d$ features
to $d'$ and $d''$ features, respectively,
is a transformer that maps its $d$ input features to $d' + d''$ features
by calling its underlying transformers in parallel.
From an API point of view, a
\texttt{FeatureUnion} takes as input a list of transformers. Calling
\texttt{fit} on the union is the same as calling \texttt{fit} independently on
each of the transformers.

@@

As an example, the snippet below illustrates how to find the
hyper-parameter settings of an SVM classifier (SVC)
that maximize its $F_1$ score,
through 10-fold cross-validation on the training set.
\begin{pythoncode}
param_grid = [
  {"kernel": ["linear"], "C": [1, 10, 100, 1000]},
  {"kernel": ["rbf"], "C": [1, 10, 100, 1000],
                      "gamma": [0.001, 0.0001]}
]

@@

In other places of the library, in particular in the feature extraction and
vectorization code for text documents, the toolkit is also designed to be
extensible. Here, estimators provide hooks for user-defined code: objects or
functions that provide a specific API can be given as arguments at vectorizer
construction time. The library then calls into this code to obtain the desired
behavior. Again, such external user code can be kept agnostic of the \sklearn
class hierarchy.
Communication proceeds by passing object of standard Python/NumPy types.

>>>>

Started in 2007, \sklearn is developed by an international team of a dozen of
core developers, mostly researchers from various fields (e.g., machine learning,
natural language processing, or neuroscience). The project also benefits
from many occasional contributors proposing small code or documentation fixes or
improvements. Development proceeds on GitHub\footnote{\url{https://github.com/scikit-learn}},
a platform which greatly facilitates this kind of
collaboration. Because of the large number of developers, emphasis is
put on keeping the project maintainable. In particular, code must follow
specific quality guidelines, such as style consistency and unit-test coverage.
Complete documentation and examples are required for all features,
and major modifications must pass code review by at least two
developers not involved in the implementation of the proposed change.

Over the past few years, the project has proved valuable to an ever increasing
user base. Its popularity can be gaged from various indicators such as hundreds
of citations in scientific publications, successes in various machine learning
challenges (e.g., Kaggle), or significant statistics derived from our
repositories and mailing list.  At the time of writing, the project is watched
by 1309 people and forked 667 times on GitHub; the mailing list receives
more than 300 mails per month; and version control logs
% ddaa494c116e3c16bf032003c5cccbed851733d2
show 183 unique contributors to the codebase.

In a previous paper \citep{pedregosa2011}, \sklearn was briefly presented and
benchmarked against several competitors. In this paper, we instead focus on an
in-depth discussion and analysis of the design choices used in building the
library. We first present in section \ref{sec:core-api} the central application
programming interface (API) and then describe, in section \ref{sec:advanced-api},
advanced API mechanisms built from the core interface. Section
\ref{sec:implementation} briefly describes the implementation.
Section \ref{sec:comparison} then
compares \emph{scikit-learn} to other major projects in terms of API. Finally,
conclusions and directions of future works are included in section
\ref{sec:conclusions}.

@@

To illustrate the initialize-\texttt{fit} sequence,
let us consider a supervised learning task using logistic regression.
Given the API defined above, solving this problem is as simple as:\footnote{
  \texttt{import} statement are omitted for brevity
  throughout this paper.}
\begin{pythoncode}
clf = LogisticRegression(penalty="l1")
clf.fit(X_train, y_train)
\end{pythoncode}
In this snippet, a \texttt{LogisticRegression} estimator is first initialized by
setting the \texttt{penalty} hyper-parameter to \texttt{"l1"}, hence enabling
$\ell_1$ regularization. Other hyper-parameters (such as \texttt{C}, which
defines the strength of the regularization) are not explicitly defined and are
thus set to the default values. Upon calling \texttt{fit}, the model is then
learned from the training arrays \texttt{X\_train} and \texttt{y\_train}.
(In this example, \texttt{fit} is called solely for its side effects.)
Since all estimators share the same interface, using a different learning algorithm is
as simple as replacing the constructor (the class name).
For example, to build a random forest on
the same training data, one would simply have to replace
\texttt{LogisticRegression(penalty="l1")} in the snippet above by \\
\texttt{RandomForestClassifier()}.

@@

Since it is common to modify or filter data before feeding it to a learning
algorithm, some estimators in the library implement a \textit{transformer}
interface which defines a \texttt{transform} method. It takes as input some new
data \texttt{X\_test} and yields as output a transformed version of
\texttt{X\_test}. Preprocessing, feature selection and dimensionality reduction
algorithms are all provided as transformers within the library.  In our example,
if one wanted to preprocess the inputs \texttt{X\_train} before fitting the
logistic regression estimator, e.g., by standardizing the feature values, one
could thus simply write:
\begin{pythoncode}
scaler = StandardScaler()
scaler.fit(X_train)
X_train = scaler.transform(X_train)
\end{pythoncode}
Of course, in practice, it is important to apply the same preprocessing to the
test data \texttt{X\_test}.
Since a \texttt{StandardScaler} stores the mean and standard deviation
that it computed for the training set,
applying it to an unseen test set \texttt{X\_test}
maps that into the appropriate region of feature space:
\begin{pythoncode}
X_test = scaler.transform(X_test)
\end{pythoncode}
Transformers also include a variety of learning algorithms, such as
dimension reduction (PCA, manifold learning), kernel approximation,
and other mappings from one feature space to another.

@@

Furthermore, every transformer allows \texttt{fit(X\_train).transform(X\_train)}
to be written as \texttt{fit\_transform(X\_train)}.
The combined \texttt{fit\_transform} method prevents repeated computations.
Depending on the transformer,
it may skip only an input validation step,
or in fact use a more efficient algorithm for the transformation.
In the same spirit, clustering algorithms provide a
\texttt{fit\_predict} method
that is equivalent to \texttt{fit} followed by \texttt{predict},
returning cluster labels assigned to the training samples.

@@

As an example, a logistic regression classifier
uses by default a one-vs.-rest scheme
for performing multiclass classification.
A different scheme can be achieved
by a meta-estimator wrapping a logistic regression estimator:
\begin{pythoncode}
ovo_lr = OneVsOneClassifier(LogisticRegression(penalty="l1"))
\end{pythoncode}
The \texttt{OneVsOneClassifier} object
\textit{clones} the LR estimator multiple times for learning,
resulting in a set of $\frac{K(K-1)}{2}$ estimator objects
for $K$-way classification,
all with the same settings,
that each perform a binary classification and vote to make the final decision.
The snippet exemplifies the importance
of separating object instantiation and actual learning.

@@

\texttt{Pipeline} objects chain multiple estimators into a
single one. This is useful since a machine learning workflow typically involves
a fixed sequence of processing steps (e.g., feature extraction,
dimensionality reduction, learning and making predictions). A sequence
of $N$ such steps can be combined into a \texttt{Pipeline}
if the first $N-1$ steps are transformers; the last can
be either a predictor or a transformer. Conceptually, fitting a pipeline to a
training set amounts to the following recursive procedure: i) when only one step
remains, call its \texttt{fit} method; ii) otherwise, \texttt{fit} the first
step, use it to \texttt{transform} the training set and \texttt{fit} the rest of
the pipeline with the transformed data. The pipeline exposes all the methods the
last estimator in the pipe exposes. That is, if the last estimator is a
predictor, the pipeline can itself be used as a predictor. If the last estimator
is a transformer, then the pipeline is itself a transformer.

\texttt{FeatureUnion} combines multiple transformers
into a single one that concatenates their outputs.
I.e., a union of transformers that map inputs having $d$ features
to $d'$ and $d''$ features, respectively,
is a transformer that maps its $d$ input features to $d' + d''$ features
by calling its underlying transformers in parallel.
In terms of the API, a
\texttt{FeatureUnion} takes as input a list of transformers. Calling
\texttt{fit} on the union is the same as calling \texttt{fit} independently on
each of the transformers.

@@

The snippet below illustrates how to find
hyper-parameter settings for an SVM classifier (SVC)
that maximize $F_1$ score
through 10-fold cross-validation on the training set.
\begin{pythoncode}
param_grid = [
  {"kernel": ["linear"], "C": [1, 10, 100, 1000]},
  {"kernel": ["rbf"], "C": [1, 10, 100, 1000],
                      "gamma": [0.001, 0.0001]}
]

@@

In other places of the library, in particular in the vectorization code
for unstructured inputs, the toolkit is also designed to be
extensible. Here, estimators provide hooks for user-defined code: objects or
functions that provide a specific API can be given as arguments at vectorizer
construction time. The library then calls into this code.
Again, such external user code can be kept agnostic of the \sklearn
class hierarchy.
Communication proceeds by passing object of standard Python/NumPy types.
